---
title: "Lab 2: Exploratory Data Analysis and Causal Model Bulding"
author: 'Team 2: Savita Chari, Denny Lehman, Tymon Silva'
date: "December 7, 2021"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, message = FALSE}
library(tidyverse)
library(dplyr)
library(lmtest)
library(sandwich)
library(stargazer)
library(magrittr)
library(sandwich)
library(gridExtra)
library(funModeling)
library(cowplot)
library(MASS)
library(ggplot2)
```
# Data

```{r load data, echo = TRUE, message=FALSE, warning=FALSE}
fire_raw <- read_csv(file = '../src/data/forestfires.csv')

# add BUI variable
BUI_less <- 0.8*((fire_raw$DMC*fire_raw$DC)/(fire_raw$DMC+(0.4*fire_raw$DC)))
BUI_great <- fire_raw$DMC-(1-((0.8*fire_raw$DC)/(fire_raw$DMC+0.4*fire_raw$DC)))*(0.92+(0.0114*fire_raw$DMC)^1.7)

fire_data <- fire_raw %>%
  mutate(
    BUI = case_when(
      DMC <= 0.4*DC ~ BUI_less,
      DMC > 0.4*DC ~ BUI_great,
    )
  )

# by adding BUI variable, we will remove DMC and DC variables that were used in
# the BUI calculation
fire_data <- fire_data %>%
  dplyr::select(wind, rain, temp, RH, FFMC, BUI, area)

```

```{r split data into EDA and Prod datasets}
# Split the data into training and testing sets
# We split the data into an EDA and a Prod dataset because we had a large enough dataset.
# We kept 30% data for EDA set and 70% for the Prod dataset.

sample_size = floor(0.7*nrow(fire_data))
set.seed(777)

# randomly split data in r
picked = sample(seq_len(nrow(fire_data)),size = sample_size)
Prod = fire_data[picked,]
EDA = fire_data[-picked,]
```

```{r EDA: Glimpse into the data by describing it, echo = TRUE}
describe(EDA)
```

```{r EDA: Glimpse into the histograms for numerical variables of interest, echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
# EDA: Glimpse into the histograms for numerical variables of interest
hist_plots <- plot_num(EDA)
```

# Analysis of Outcome Variable : area

Area is very important for our study because it is our outcome variable. We want to find out what factors cause area to burn in forest fires.
From the graph above, area seems to have a heavy left tail. We should use log transformation on area field to see if it helps improve a linear relationship with our predictor variables.

```{r EDA: Observations with area = 0,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
cat ("Total observations in the dataset: ", nrow(EDA))
cat("\nNumber of Observations where area is zero:" ,nrow(subset(EDA,area == 0)))
```
```{r EDA: Plot the histogram of the Distribution of Burnt Area,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
# Histogram of Non Log Transformed area
hist_of_area_dist <- EDA %>% 
ggplot() + aes(x = area) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) +
labs(
x = "Area in hectare ", y="Occurrence",
title = '               Burnt area Distribution \n \nNon-Log Transformed',
caption = "Figure 1"
)


# Perform Log transformation
EDA$logarea <- log(EDA$area + 1)

# Histogram of Log Transformed area
hist_of_logarea_dist <- EDA %>% 
ggplot() + aes(x = logarea) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) +
labs(
x = "Area in hectare ", y="Occurrence",
title = '\n\nLog Transformed ',
caption = "Figure 2"
)

# Plot both the histograms
plot_grid(hist_of_area_dist,hist_of_logarea_dist)

```

# Observation:
The graphs above has a very heavy left tail and log transformation helped improve the distribution to look more normal. However, it is still skewed right. 

#  Analyzing of Linear Relationship of Predictors Variables with Outcome

In our causal diagram we have identified four predictor variables 
1. rain - outside rain in mm/m2 : 0.0 to 6.4
2. wind - wind speed in km/h: 0.40 to 9.40
3. FFMC - FFMC index from the FWI system: 18.7 to 96.20
4. BUI - potential heat release in heavier fuels (total amount of fuel available for combustion): 0.0 to infinity

```{r plotting predictors vs log transformed outcome variable}
a <- ggplot(EDA, aes(x = wind, y = log(area+1))) +
  geom_point() +
  geom_smooth() +
  ggtitle("wind vs log(area+1)")

b <- ggplot(EDA, aes(x = rain, y = log(area+1))) +
  geom_point() +
  geom_smooth() +
  ggtitle("rain vs log(area+1)")

c <- ggplot(EDA, aes(x = FFMC, y = log(area+1))) +
  geom_point() +
  geom_smooth() +
  ggtitle("FFMC vs log(area+1)")

d <- ggplot(EDA, aes(x = BUI, y = log(area+1))) +
  geom_point() +
  geom_smooth() +
  ggtitle("BUI vs log(area+1)")

grid.arrange(a, b, c, d, nrow = 2)
```
# Observation:
Rain and BUI variables have room for improvement. Will try log transforms to see if it helps improve linearity.

```{r plotting log trasnformed rain and BUI predictors vs outcome variable}
a <- ggplot(EDA, aes(x = rain, y = log(area+1))) +
  geom_point() +
  geom_smooth() +
  ggtitle("rain vs log(area+1)")

b <- ggplot(EDA, aes(x = log(rain+1), y = log(area+1))) +
  geom_point() +
  geom_smooth() +
  ggtitle("log(rain) vs log(area+1)")

c <- ggplot(EDA, aes(x = BUI, y = log(area+1))) +
  geom_point() +
  geom_smooth() +
  ggtitle("BUI vs log(area+1)")

d <- ggplot(EDA, aes(x = log(BUI), y = log(area+1))) +
  geom_point() +
  geom_smooth() +
  ggtitle("log(BUI) vs log(area+1)")

grid.arrange(a, b, c, d, nrow = 2)
```
## Conclusion
rain predictor did not benefit from a log transform. However, log(BUI) seems to have slightly improved linear relationship with log(area+1).

# Analysis of rain

```{r last resort in fixing rain variable}

# all transformations failed to improve the plot of rain and log(area+1)/
# so, we will convert our rain variable to binary, is not raining (0) or is raining (1).

Prod <- Prod %>%
  mutate(
    rain_binary = case_when(
      rain > 0 ~ 1,
      rain == 0 ~ 0,
      )
  )

EDA <- EDA %>%
  mutate(
    rain_binary = case_when(
      rain > 0 ~ 1,
      rain == 0 ~ 0,
      )
  )

```

# Observation: wind and FFMC 

```{r EDA: Check the distribution of column wind }
hist_of_wind_dist <- EDA %>% 
ggplot() + aes(x = wind ) +
geom_histogram( bins=20, fill="#C8E6C9", color="black", alpha=0.9) +
labs(
x = "wind speed in km/h", y="Occurrence",
title = 'Distribution graphs',
subtitle = 'Wind occurrence'
)

hist_of_FFMC_dist <- EDA %>% 
ggplot() + aes(x = log(FFMC) ) +
geom_histogram( bins=20, fill="#FFCCBC", color="black", alpha=0.9) +
labs(
x = "FFMC code", y="Occurrence",
subtitle = 'Fine Fuel Moisture Code'
)



par(mfrow=c(2, 1))
    
plot_grid(hist_of_wind_dist,hist_of_FFMC_dist)

```
# Obseration
Fairly well distributed. No transformation needed.

```{r Multi-Variable relationship, echo = FALSE, message=FALSE, warning=FALSE}
scatterplotMatrix(~ log(area+1) + wind + FFMC +log(BUI),  data = EDA,
                  diagonal = FALSE,             # Remove kernel density estimates
                  regLine = list(col = "green", # Linear regression line color
                                 lwd = 3),      # Linear regression line width
                  smooth = list(col.smooth = "red",   # Non-parametric mean color
                                col.spread = "blue")) # Non-parametric variance color

```

# Building Our Linear Models 
```{r model Building: }
# our primary model used to answer our Research Question
model1 <- lm(log(area+1) ~ wind, data = Prod)
coeftest(model1, vcov=vcovHAC)

# a model that uses all available variables
model2 <- lm(log(area+1) ~ wind + rain_binary, data = Prod)
coeftest(model2, vcov=vcovHAC)

# a model that uses only the direct atmospheric metrics
model3 <- lm(log(area+1) ~ wind + rain_binary + FFMC + log(BUI), data = Prod)
coeftest(model3, vcov=vcovHAC)

```


```{r print model estimates, warning = FALSE}
# while you are writing you code, you can use `type = 'text'` to print to the console
stargazer(
  model1,
  model2,
  model3,
  type = 'text', header = FALSE,
  star.cutoffs = c(0.05, 0.01, 0.001) # the default isn't in line with w203
)
```

```{r Plotting the model}
anova(model1, model2, model3, test="F")
```
