---
title: "Lab 2: Exploratory Data Analysis and Causal Model Bulding"
author: 'Team 2: Savita Chari, Denny Lehman, Tymon Silva'
date: "December 7, 2021"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, message = FALSE}
library(tidyverse)
library(dplyr)
library(lmtest)
library(sandwich)
library(stargazer)
library(magrittr)
library(sandwich)
library(gridExtra)
library(funModeling)
library(cowplot)
library(MASS)
library(ggplot2)
```
# Data

```{r load data, echo = TRUE, message=FALSE, warning=FALSE}
fire_raw <- read_csv(file = '../src/data/forestfires.csv')

# add BUI variable
BUI_less <- 0.8*((fire_raw$DMC*fire_raw$DC)/(fire_raw$DMC+(0.4*fire_raw$DC)))
BUI_great <- fire_raw$DMC-(1-((0.8*fire_raw$DC)/(fire_raw$DMC+0.4*fire_raw$DC)))*(0.92+(0.0114*fire_raw$DMC)^1.7)

fire_data <- fire_raw %>%
  mutate(
    BUI = case_when(
      DMC <= 0.4*DC ~ BUI_less,
      DMC > 0.4*DC ~ BUI_great,
    )
  )

# by adding BUI variable, we will remove DMC and DC variables that were used in
# the BUI calculation
fire_data <- fire_data %>%
  dplyr::select(wind, rain, temp, RH, FFMC, BUI, area)

```

```{r split data into EDA and Prod datasets}
# Split the data into training and testing sets
# We split the data into an EDA and a Prod dataset because we had a large enough dataset.
# We kept 30% data for EDA set and 70% for the Prod dataset.

sample_size = floor(0.7*nrow(fire_data))
set.seed(777)

# randomly split data in r
picked = sample(seq_len(nrow(fire_data)),size = sample_size)
Prod = fire_data[picked,]
EDA = fire_data[-picked,]
```

```{r EDA: Glimpse into the data by describing it, echo = TRUE}
describe(EDA)
```

```{r EDA: Glimpse into the histograms for numerical variables of interest, echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
# EDA: Glimpse into the histograms for numerical variables of interest
hist_plots <- plot_num(EDA)
```

# Analysis of Outcome Variable : area

Area is very important for our study because it is our outcome variable. We want to find out what factors cause area to burn in forest fires.
From the graph above, area seems to have a heavy left tail. We should use log transformation on area field to see if it helps improve a linear relationship with our predictor variables.

```{r EDA: Observations with area = 0,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
cat ("Total observations in the dataset: ", nrow(EDA))
cat("\nNumber of Observations where area is zero:" ,nrow(subset(EDA,area == 0)))
```
```{r EDA: Plot the histogram of the Distribution of Burnt Area,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
# Histogram of Non Log Transformed area
hist_of_area_dist <- EDA %>% 
ggplot() + aes(x = area) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) +
labs(
x = "Area in hectare ", y="Occurrence",
title = '               Burnt area Distribution \n \nNon-Log Transformed',
caption = "Figure 1"
)


# Perform Log transformation
EDA$logarea <- log(EDA$area + 1)

# Histogram of Log Transformed area
hist_of_logarea_dist <- EDA %>% 
ggplot() + aes(x = logarea) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) +
labs(
x = "Area in hectare ", y="Occurrence",
title = '\n\nLog Transformed ',
caption = "Figure 2"
)

# Plot both the histograms
plot_grid(hist_of_area_dist,hist_of_logarea_dist)

```

# Observation:
The graphs above has a very heavy left tail and log transformation helped improve the distribution to look more normal. However, it is still skewed right. 

```{r EDA: Plot the histogram of the Distribution of Burnt Area,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }

set.seed(1234)
hist_of_area_dist <- EDA %>% 
ggplot() + aes(x = area) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) + 
labs(
x = "Area in hectare ", y="Occurrence",
title = 'Values > 0 Burnt area Distribution',
subtitle = 'Non-Log Transformed',
caption = "Figure 3"
) 

#hist_of_area_dist

# Perform Log transformation
EDA$logarea <- log(EDA$area + 1)


hist_of_logarea_dist <- EDA %>% 
ggplot() + aes(x = logarea) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) + 
  
  geom_vline(aes(xintercept = mean(logarea)), color = "#000000", size = 1.25) +
  geom_vline(aes(xintercept = mean(logarea) + sd(logarea)), color = "#000000", size = 1, linetype = "dashed") +
  geom_vline(aes(xintercept = mean(logarea) - sd(logarea)), color = "#000000", size = 1, linetype = "dashed")+
labs(
x = "Area in hectare ", y="Occurrence",
subtitle = '\n\nLog Transformed (with intercept)',
caption = "Figure 4"
)

    
plot_grid(hist_of_area_dist,hist_of_logarea_dist)

```
# Observation: Area is our outcome variable. After performing the log tansformation and filtering out the zero values of the area, the distribution of our outcome variable 'area' looks normal (Figure 4)


#  Analysing of Predictors Variables

In our causal diagram we have identified four predictor variables 
1. rain - outside rain in mm/m2 : 0.0 to 6.4
2. wind - wind speed in km/h: 0.40 to 9.40
3. FFMC - FFMC index from the FWI system: 18.7 to 96.20
4. BUI - 

# Analysis of rain

```{r EDA: Histogram to see if rain has normal distribution,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE}
hist_of_rain_dist <- EDA %>% 
ggplot() + aes(x = rain ) +
geom_histogram( bins=20, fill="#B0DEEA", color="black", alpha=0.9) +
labs(
x = "rainfall in mm/m2", y="Occurrence",
title = 'Distribution graphs',
subtitle = 'rain occurrence'
)

# Log trnsformation on rain

hist_of_lograin_dist <- EDA %>% 
ggplot() + aes(x = log(rain) ) +
geom_histogram( bins=20, fill="#B0DEEA", color="black", alpha=0.9) +
labs(
x = "rainfall in mm/m2", y="Occurrence",
title = 'Log Transformed rain',
subtitle = 'rain occurrence',

)
plot_grid(hist_of_rain_dist,hist_of_lograin_dist)

```
# Observation - rain
The above histograms shows thhat rain did not benifit from log transformation
Let's draw a box plot to observe how does the data look!

```{r EDA: Box plotting predictors vs log transformed outcome variable , echo = FALSE, message=FALSE, warning=FALSE}
#  Box plots is a great way to quickly examin our predictor variables
a <- ggplot(EDA, aes(x = wind, y = logarea)) +
  geom_boxplot() +
  geom_smooth() +
  ggtitle("wind vs log(area)")

b <- ggplot(EDA, aes(x = rain, y = logarea)) +
  geom_boxplot() +
  geom_smooth() +
  ggtitle("rain vs log(area)")

c <- ggplot(EDA, aes(x = FFMC, y = logarea)) +
  geom_boxplot() +
  geom_smooth() +
  ggtitle("FFMC vs log(area)")



grid.arrange(a, b, c, nrow = 2)
```
# Observation: rain (continued)
Most of the observations do not have any data for rain.
All of the very sparce existing rain data is outlier, this is the reason log transformation did not make any difference to rain.
In this state, rain will not help in answering our causal question
# Further Transformation on rain
We will convert rain to a binary (True/False) field 

```{r Transform rain field to a binary field.}
# For all the observation that has some value in rain column, the rain_binary will be True, else it will be False
EDA <- EDA %>%
  mutate(
    rain_binary = case_when(
      rain > 0 ~ 1,
      rain == 0 ~ 0,
      )
  )

```

# Observation: wind and FFMC 

```{r EDA: Check the distribution of column wind }
hist_of_wind_dist <- EDA %>% 
ggplot() + aes(x = wind ) +
geom_histogram( bins=20, fill="#C8E6C9", color="black", alpha=0.9) +
labs(
x = "wind speed in km/h", y="Occurrence",
title = 'Distribution graphs',
subtitle = 'Wind occurrence'
)

hist_of_FFMC_dist <- EDA %>% 
ggplot() + aes(x = log(FFMC) ) +
geom_histogram( bins=20, fill="#FFCCBC", color="black", alpha=0.9) +
labs(
x = "FFMC code", y="Occurrence",
subtitle = 'Fine Fuel Moisture Code'
)



par(mfrow=c(2, 1))
    
plot_grid(hist_of_wind_dist,hist_of_FFMC_dist)

```
# Obseration
Fairly well distributed. No transformation neede.

```{r Multi-Variable relationship, echo = FALSE, message=FALSE, warning=FALSE}
scatterplotMatrix(~ logarea + wind + FFMC ,  data = EDA,
                  diagonal = FALSE,             # Remove kernel density estimates
                  regLine = list(col = "green", # Linear regression line color
                                 lwd = 3),      # Linear regression line width
                  smooth = list(col.smooth = "red",   # Non-parametric mean color
                                col.spread = "blue")) # Non-parametric variance color

```
# Building model model_SAV_long on data without area = 0

```{r model Building: model_SAV_long }
model_SAV_long <- lm(logarea ~ rain_binary + FFMC + wind , data = EDA)
coeftest(model_SAV_long, vcov=vcovHAC)
```

# Building model model_SAV_long on data with area = 0

```{r model Building: model_SAV_long }
model_SAV_long <- lm(logarea ~ rain_binary + FFMC + wind , data = EDA)
coeftest(model_SAV_long, vcov=vcovHAC)
```
# Building model model_SAV_long on data with area = 0 with temp as a predictor variable

```{r model Building: model_SAV_long }
model_SAV_long <- lm(logarea ~ rain_binary + FFMC + wind + temp, data = EDA)
coeftest(model_SAV_long, vcov=vcovHAC)
```
# Building model model1 on data without area = 0
```{r model building : model1}
model1 <- lm(log(area+1) ~ wind + rain_binary + temp + FFMC + log(DMC) + DC, data = EDA)
coeftest(model1, vcov = vcovHC)
```




# Building model model1 on data with area = 0
```{r model building : model1}
model1 <- lm(log(area+1) ~ wind + rain_binary + temp + FFMC + log(DMC) + DC, data = EDA)
coeftest(model1, vcov = vcovHC)
```

```{r model building: model2}

model2 <- lm(log(area+1) ~ wind + rain_binary + temp + FFMC + log(DMC) + DC + RH, data = EDA)

coeftest(model2, vcov = vcovHC)
```



```{r model building: model3}
model3 <- lm(log(area+1) ~ wind + rain_binary + temp + RH, data = EDA)
coeftest(model3, vcov = vcovHC)
```


```{r Plotting the model}
set.seed(5600)
shapiro.test(sample(model_SAV_long$residuals, size = 5000,replace=TRUE))
plot(model1)

```













