---
title: "Lab 2: Exploratory Data Analysis and Causal Model Bulding"
author: 'Team 2: Savita Chari, Denny Lehman, Tymon Silva'
date: "December 7, 2021"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, message = FALSE}
library(tidyverse)
 
library(lmtest)
library(sandwich)
library(stargazer)
library(magrittr)
library(sandwich)
library(gridExtra)
library(funModeling)
library(cowplot)
library(MASS)
library(ggplot2)
```
# Data

```{r load data, echo = TRUE, message=FALSE, warning=FALSE}
eda_data <- read_csv(file = '../src/data/forestfires.csv')
```


```{r EDA: Glimpse into the data by describing it, echo = TRUE}
describe(eda_data)
```

```{r EDA: Glimpse into the histograms for numerical variables of interest, echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
# EDA: Glimpse into the histograms for numerical variables of interest
p2 <- plot_num(eda_data[5:13])


```

# Analysis of Outcome Variable : area

Area is very important for our study because it is our outcome variable. We want to find out what factors cause area to burn in forest fires.
From the graph above, area seems to have a heavy left tail. We should use log transformation on area field to see if the IID improves.

```{r EDA: Observations with area = 0,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
cat ("Total observations in the dataset: ", nrow(eda_data))
cat("\nNumber of Observations where area is zero:" ,nrow(subset(eda_data,area == 0)))
```
```{r EDA: Plot the histogram of the Distribution of Burnt Area,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
# Histogram of Non Log Transformed area
hist_of_area_dist <- eda_data %>% 
ggplot() + aes(x = area) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) +
labs(
x = "Area in hectare ", y="Occurrence",
title = '               Burnt area Distribution \n \nNon-Log Transformed',
caption = "Figure 1"
)


# Perform Log transformation
eda_data$logarea <- log(eda_data$area + 1)

# Histogram of Log Transformed area
hist_of_logarea_dist <- eda_data %>% 
ggplot() + aes(x = logarea) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) +
labs(
x = "Area in hectare ", y="Occurrence",
title = '\n\nLog Transformed ',
caption = "Figure 2"
)

# Plot both the histograms
plot_grid(hist_of_area_dist,hist_of_logarea_dist)

```

# Observation:
The graphs above has a very heavy left tail and log transformation did not fix it. As observed earlier,  a lot of observations have area = 0. Which essentially means that these observations did not cause any measurable area to burn. For our analysis such rows are not important. Nearly 48% of the rows have zero value in area field. So  we should remove these rows and see if that improves the distribution.



```{r Perform log transformation on area field. Pick only the observations with some data in the area column}
eda_sub_data <- eda_data %>%  filter(area > 0)

#view(eda_sub_data)
```


```{r EDA: Observations with area = 0,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }
cat ("Total observations in the dataset: ", nrow(eda_sub_data))
cat("\nNumber of Observations where area is zero:" ,nrow(subset(eda_sub_data,area == 0)))
```
# Repeat the analysis done abve on the sub set of data where the area (of burn) has a value > 0

```{r Describe the sub set of data}
summary(eda_sub_data)

```

```{r,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE}
plot_num(eda_sub_data[5:13])

```


```{r EDA: Plot the histogram of the Distribution of Burnt Area,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE }

set.seed(1234)
hist_of_area_dist <- eda_sub_data %>% 
ggplot() + aes(x = area) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) + 
labs(
x = "Area in hectare ", y="Occurrence",
title = 'Values > 0 Burnt area Distribution',
subtitle = 'Non-Log Transformed',
caption = "Figure 3"
) 

#hist_of_area_dist

# Perform Log transformation
eda_sub_data$logarea <- log(eda_sub_data$area)


hist_of_logarea_dist <- eda_sub_data %>% 
ggplot() + aes(x = logarea) +
geom_histogram( bins=20, fill="cornsilk", color="black", alpha=0.9) + 
  
  geom_vline(aes(xintercept = mean(logarea)), color = "#000000", size = 1.25) +
  geom_vline(aes(xintercept = mean(logarea) + sd(logarea)), color = "#000000", size = 1, linetype = "dashed") +
  geom_vline(aes(xintercept = mean(logarea) - sd(logarea)), color = "#000000", size = 1, linetype = "dashed")+
labs(
x = "Area in hectare ", y="Occurrence",
subtitle = '\n\nLog Transformed (with intercept)',
caption = "Figure 4"
)

    
plot_grid(hist_of_area_dist,hist_of_logarea_dist)

```
# Observation: Area is our outcome variable. After performing the log tansformation and filtering out the zero values of the area, the distribution of our outcome variable 'area' looks normal (Figure 4)


#  Analysing of Predictors Variables

In our causal diagram we have identified three predictor variables 
1. rain - outside rain in mm/m2 : 0.0 to 6.4
2. wind - wind speed in km/h: 0.40 to 9.40
3. FFMC - FFMC index from the FWI system: 18.7 to 96.20


# Analysis of rain

```{r EDA: Histogram to see if rain has normal distribution,  echo = FALSE, message=FALSE, warning=FALSE, show_col_types = FALSE}
hist_of_rain_dist <- eda_sub_data %>% 
ggplot() + aes(x = rain ) +
geom_histogram( bins=20, fill="#B0DEEA", color="black", alpha=0.9) +
labs(
x = "rainfall in mm/m2", y="Occurrence",
title = 'Distribution graphs',
subtitle = 'rain occurrence'
)

# Log trnsformation on rain

hist_of_lograin_dist <- eda_sub_data %>% 
ggplot() + aes(x = log(rain) ) +
geom_histogram( bins=20, fill="#B0DEEA", color="black", alpha=0.9) +
labs(
x = "rainfall in mm/m2", y="Occurrence",
title = 'Log Transformed rain',
subtitle = 'rain occurrence',

)
plot_grid(hist_of_rain_dist,hist_of_lograin_dist)

```
# Observation - rain
The above histograms shows thhat rain did not benifit from log transformation
Let's draw a box plot to observe how does the data look!

```{r EDA: Box plotting predictors vs log transformed outcome variable , echo = FALSE, message=FALSE, warning=FALSE}
#  Box plots is a great way to quickly examin our predictor variables
a <- ggplot(eda_sub_data, aes(x = wind, y = logarea)) +
  geom_boxplot() +
  geom_smooth() +
  ggtitle("wind vs log(area)")

b <- ggplot(eda_sub_data, aes(x = rain, y = logarea)) +
  geom_boxplot() +
  geom_smooth() +
  ggtitle("rain vs log(area)")

c <- ggplot(eda_sub_data, aes(x = FFMC, y = logarea)) +
  geom_boxplot() +
  geom_smooth() +
  ggtitle("FFMC vs log(area)")



grid.arrange(a, b, c, nrow = 2)
```
# Observation: rain (continued)
Most of the observations do not have any data for rain.
All of the very sparce existing rain data is outlier, this is the reason log transformation did not make any difference to rain.
In this state, rain will not help in answering our causal question
# Further Transformation on rain
We will convert rain to a binary (True/False) field 

```{r Transform rain field to a binary field.}
# For all the observation that has some value in rain column, the rain_binary will be True, else it will be False
eda_sub_data <- eda_sub_data %>%
  mutate(
    rain_binary = case_when(
      rain > 0 ~ 1,
      rain == 0 ~ 0,
      )
  )

```

# Observation: wind and FFMC 

```{r EDA: Check the distribution of column wind }
hist_of_wind_dist <- eda_sub_data %>% 
ggplot() + aes(x = wind ) +
geom_histogram( bins=20, fill="#C8E6C9", color="black", alpha=0.9) +
labs(
x = "wind speed in km/h", y="Occurrence",
title = 'Distribution graphs',
subtitle = 'Wind occurrence'
)

hist_of_FFMC_dist <- eda_sub_data %>% 
ggplot() + aes(x = log(FFMC) ) +
geom_histogram( bins=20, fill="#FFCCBC", color="black", alpha=0.9) +
labs(
x = "FFMC code", y="Occurrence",
subtitle = 'Fine Fuel Moisture Code'
)



par(mfrow=c(2, 1))
    
plot_grid(hist_of_wind_dist,hist_of_FFMC_dist)

```
# Obseration
Fairly well distributed. No transformation neede.

```{r Multi-Variable relationship, echo = FALSE, message=FALSE, warning=FALSE}
scatterplotMatrix(~ logarea + wind + FFMC +temp ,  data = eda_sub_data,
                  diagonal = FALSE,             # Remove kernel density estimates
                  regLine = list(col = "green", # Linear regression line color
                                 lwd = 3),      # Linear regression line width
                  smooth = list(col.smooth = "red",   # Non-parametric mean color
                                col.spread = "blue")) # Non-parametric variance color

```
# Building model model_SAV_long on data without area = 0

```{r model Building: model_SAV_long }
model_SAV_long <- lm(logarea ~ rain_binary + FFMC + wind , data = eda_sub_data)
coeftest(model_SAV_long, vcov=vcovHAC)
```

# Building model model_SAV_long on data with area = 0

```{r model Building: model_SAV_long }
model_SAV_long <- lm(logarea ~ rain_binary + FFMC + wind , data = eda_data)
coeftest(model_SAV_long, vcov=vcovHAC)
```
# Building model model_SAV_long on data with area = 0 with temp as a predictor variable

```{r model Building: model_SAV_long }
model_SAV_long <- lm(logarea ~ rain_binary + FFMC + wind + temp, data = eda_data)
coeftest(model_SAV_long, vcov=vcovHAC)
```
# Building model model1 on data without area = 0
```{r model building : model1}
model1 <- lm(log(area+1) ~ wind + rain_binary + temp + FFMC + log(DMC) + DC, data = eda_sub_data)
coeftest(model1, vcov = vcovHC)
```




# Building model model1 on data with area = 0
```{r model building : model1}
model1 <- lm(log(area+1) ~ wind + rain_binary + temp + FFMC + log(DMC) + DC, data = eda_data)
coeftest(model1, vcov = vcovHC)
```

```{r model building: model2}

model2 <- lm(log(area+1) ~ wind + rain_binary + temp + FFMC + log(DMC) + DC + RH, data = eda_data)

coeftest(model2, vcov = vcovHC)
```



```{r model building: model3}
model3 <- lm(log(area+1) ~ wind + rain_binary + temp + RH, data = eda_data)
coeftest(model3, vcov = vcovHC)
```


```{r Plotting the model}
set.seed(5600)
shapiro.test(sample(model_SAV_long$residuals, size = 5000,replace=TRUE))
plot(model1)

```













